The Evolution of Phenotypically Invariant Genetic Networks

### 2017-09-13 kalman_walk_paper.tex
added an informal description of the kalman decomposition to the main text. Edited metabolic example and recombination section.

%Define $V(\tau)$ as the parameterized change of coordinates matrix that preserves $B$ and $C$, with $\tau$ a vector of free parameters. The set of \emph{all} phenotypically invariant (minimal) gene networks is, 
 % \begin{align}
 %   A(\tau) = V(\tau) A(0) V^{-1}(\tau) ,
 % \end{align}
%and a \emph{Linear Evolutionary System} is, 
%\plr{why is this "evolutionary"? It is the thing that we're thinking about evolution of, but that doesn't make it evolutionary.}
Therefore the set of all phenotypically identical (and minimal) systems can be parameterized as, $\mathcal{S}_{\min}(V) = (VA_0V^{-1}, VB_0, C_0V )$,
  %  \begin{align} \label{eq:min_param}
  %    \mathcal{S}(V) := \left\{ \begin{array}{ll} V\dot{\kappa}(t) &= VA V^{-1} \kappa(t) + VB u(t) \\ \phi(t) &= C V^{-1} \kappa(t) \end{array} \right.
  %\end{align}
  where $V$ is any invertible matrix and it's elements are free parameters.

@@

More generally,
we denote by $\calA_n(\Sys_0)$ the set of all $n$-dimensional systems equivalent to $\Sys_0$:
\begin{equation} \label{eqn:equivalence}
  \begin{aligned}
    \calA_n(\Sys_0) 
      &= \left\{
        (A,B,C) : C e^{At} B = C_0 e^{A_0 t} B_0 \; \text{for}\; t \ge 0 
      \right\}  \\
      &= \left\{
        (A,B,C) : C A^r B = C_0 A_0^r B_0 \; \text{for}\; 1 \le r \le n-1 
      \right\} .
  \end{aligned}
\end{equation}
Equivalence of the two characterizations follows from the Cayley-Hamilton theorem.
\plr{To make sense of this we need to say how $C$ and $B$ change with $n$.}
Usually, the dimension $n$ and the reference system $A_0$ is implicit and we write only $\calA$.
Further, we refer to the set of all linear systems with impulse response $h(t)$, regardless of dimension, to be $\allS$,
  \begin{align}
    \allS := \bigcup_{n=\min}^{\infty} \calA_n(\Sys_0)
  \end{align}
  \js{what do you like better for $\allS$? $\mathcal{A}$, $\mathcal{A}_{\infty}$, or $\mathcal{K}_{h(t)}$, or something else -- maybe a modification of $\Sys$? Using a script of ``A'' was more intuitive when we were leavng $B$ and $C$ invariant.}
Two systems, even if in different dimensions, will have identical phenotypes if they are in $\allS$. Further, there is no unique triple $(A,B,C)$ for the mapping $h: u \mapsto \phi$. This implies that the gene regulatory network architecture per phenotype/environment pair is never unique -- that is under these assumptions, \emph{all} gene regulatory networks can drift/rewire, and still conserve the phenotype. This set can be precisley defined and completely parameterized using the \emph{Kalman decomposition} (Appendix \ref{apx:kalman}).

@@

A large, adapted population will move around in $\allS$ under environmental and selective stasis. In the minimal dimension neutral evolution proceeds as a stochastic coordinate transformation on the elements in $\Sys_{\min}$. Even if there is a conservation of phenotype through evolutionary time, the molecular mechanism underpinning the phenotype is not constrained and likely will not be conserved.

@@

%\plr{This is not part of the definition of $\mathcal{I}$:}
%\js{should we delete this example ($A(0) \times A(2)$ hybrids)? I like how clean it is -- but I'm not sure it's that important in this context. It's a good proof of principle though.}
%\jss{$F_1$s created by crossing phenotypically equivalent oscillators $A(0)$ and $A(2)$ have a phenotype of $\phi_{F_{1}}(t) = e^{t}$, 
%in contrast to both parents, who have $\phi_{P_{1}}(t) = \phi_{P_{2}}(t) = \sin(t) + \cos(t)$. 
%The hybrid phenotype is significantly different (it does not oscillate and increases infinitely) despite the phenotypic equivalence of the parents.
%%\plr{don't need these to be big equations.}
%%\begin{align*}
%%  \fit \left(\phi_{P_{1}}\right) = \fit \left( \phi_{P_{2}} \right) &= 1 \\
%%  \fit\left(\phi_{F_{1}}\right) &= 0 \\
%%  \mathcal{I} &= 0
%%\end{align*}
%
%%Thus if populations $1$ and $2$ are homogeneous $A(0)$ and $A(2)$, respectively, we say that they are completely incompatible as $\mathcal{I} = 0$.
%
%  \begin{figure}[H]
%    \centering
%    \includegraphics[width=0.5\textwidth, height=0.25\paperheight]{expF1}
%    \caption{$F_1$ hybrid (orange) and parental (blue) phenotypic oscillator dynamics for an $A(0)$ by $A(2)$ cross. The hybrid fails to oscillate and exhibits qualitatively different dynamics.}
%  \end{figure}
%  }
%%     \begin{example}[F1 Reproductive Incompatibility in an Oscillating Gene Network]
%%       DMI examples...
%%       \begin{figure}[H]
%%         \includegraphics[width=0.5\textwidth]{A2_A3_osci_F1_hyb}
%%       \end{figure}
%%     \end{example}
%
%%     \begin{example}[Not all Networks can Host Incompatibilities]
%%       convex sets cant have DMIs
%
%%       \begin{align*}
%%         h(t) = 2 e^{- \theta t}
%%       \end{align*}
%%       Any non-minimal system with rows summing to $\theta$ is PI. Further, these systems are closed under averaging (mating) and row swapping (meiosis), leaving all hybrids optimally fit. The set of gene matrices is affine and therefore convex.  
%%     \end{example}
%
%
%\plri{Missing: a statement about when there are neutral directions, and how many.}
%
%\plr{I think we should first put the section about quantitative drift in, and then the following section that tries to get at real values of parameters.}
%
%
 \begin{example}[All Phenotypically Equivalent Cell Cycle Control Networks] \label{ex:all_osc}
    The set of all two-gene regulatory networks phenotypically equivalent to the cell cycle control network in \ref{ex:oscillator}, where only $A$ can vary, is given by
    %\begin{align*}
    %  A(\tau) &= \begin{bmatrix} 1 & 0 \\ \tau & 1-\tau \end{bmatrix} \begin{bmatrix} 0 & 1 \\ -1 & 0 \end{bmatrix} \begin{bmatrix} 1 & 0 \\ \frac{\tau}{\tau-1} & \frac{-1}{\tau-1} \end{bmatrix}  \qquad \forall \ \tau \neq 1 %\\
    %\end{align*}
    \begin{align*}
      A(\tau) &= \frac{1}{\tau-1} \begin{bmatrix} \tau & -1 \\ 2 \tau(\tau - 1) + 1 &  -\tau \end{bmatrix} \ \text{for} \ \tau \neq 1
    \end{align*}
    \begin{figure}[H]
    \centering
% Nice tikz!!!
\begin{tikzpicture}
\begin{scope}[every node/.style={circle,thick,draw}]
  \node (A) at (0,0) {$\kappa_{1}$};
    \node[dashed] (B) at (4,0) {$\kappa_{2}$};
    \node[shape=rectangle] (U) at (2,2) {input};
    \node[shape=rectangle] (y) at (2,-2) {output};
\end{scope}

@@

\end{tikzpicture}
    \caption{The set of all phenotype-equivalent cell cycle control networks, $A(\tau)$.
      $A(\tau) = \left\{ \calA_2 : B = B_0,\ C = C_0\right\}$.}
    \end{figure}
 % \end{example}

@@

   Despite the phenotypic equivalence of all instantiations of $A(\tau)$, the kryptotypes, vary as a function of $\tau$. 

@@

    \caption{Gene-1 dynamics (blue) for both systems $A(0)$ and $A(2)$ are identical, however, $A(0)$ gene-2 dynamics (orange) differ from $A(2)$ (green). Both gene-1 dynamics are given by $\kappa_{1} = \sin t + \cos t$, and gene-2 by $\kappa_{2} = \cos t - \sin t$ ($A(0)$) and $\kappa'_{2} = \cos t + 3 \sin t$ ($A(2)$).}

@@

%\paragraph*{Hybrid phenotypic breakdown between oscillators}
%For example, if the two phenotypically equivalent systems $A(0)$ and $A(2)$ from Example \ref{ex:all_osc} hybridize, their offspring's phenotype will fail to oscillate completely. 
  \begin{figure}[H] 
    \centering
    \includegraphics[width=0.5\textwidth, height=0.125\paperheight]{expF1}
    \caption{\textbf{Hybrid phenotypic breakdown} $F_1$ hybrid (orange) and parental (blue) phenotypic oscillator dynamics for an $A(0)$ by $A(2)$ cross from Example \ref{ex:all_osc}. The $F_1$ hybrid
%$\left(A_{F_1}\!\!=\!\!\tfrac{A\!(\!0\!) \! + \! A\!(\!2\!)}{2}\right)$
gene expression fails to oscillate and instead grows exponentialy,
exhibiting qualitatively different dynamics ($\phi_{F_0}\! = \! \sin t \! + \! \cos t$, however $\phi_{F_1} \! = \! \exp t$).
    }
    \label{fig:expF1}
  \end{figure}
%
%\paragraph{Sexual reproduction and speciation}
%A diploid organism contains two gene network copies: one from each parent. As such, a diploid's system dynamics is computed by averaging the coefficients of both systems. To reproduce, each organism passes on a recombined haploid gene network to its offspring. Haploid gene networks swap system matrix rows randomly between its own two networks.
%Therefore the phenotypic dynamics typical of a first generation ($F_{1}$) cross between two allopatric populations will be determined simply by averaging it's two parental genomes. However, in second generation hybrid ($F_2$) crosses, first new haploid systems will be formed by recombination -- in the process shuffling and combining regulatory coefficients from allopatric populations -- and then brought together to form a diploid.
%
%
  \begin{example}[Hybrid Incompatibility in an Oscillating Gene Network] \label{ex:hybrid_osc}

Here we compare the phenotypes for F2 hybrids formed by crossing oscillators $A(2)$ with $A(2.01)$, $A(2.1)$, and $A(2.5)$ ($B$ and $C$ are the same as above). 
Each $A$ is phenotypically identical ($\phi(t) = \sin t + \cos t$), however some of the hybrids exhibit markedly different dynamics.

@@

\begin{figure}[H]
  \centering
  \begin{tabular}{ccc}
    \includegraphics[width=0.25\textwidth, height=0.125\paperheight]{F1_comparison} & 
    \includegraphics[width=0.25\textwidth, height=0.125\paperheight]{F2s_comparison2} &
    \includegraphics[width=0.25\textwidth, height=0.125\paperheight]{figures/quartic_quadratic2}
  \end{tabular}
  \caption{
    $F_1$ (left) and $F_2$ (right) hybrids crossing $A(2)$ with $A(2.01)$ (top), $A(2.1)$ (middle), and $A(2.5)$ (bottom).
    Note the difference in scale on the $y$-axis. F2 hybrids display more phenotypic divergence than F1s, on average. Further, some F2s completely fail to oscillate, as seen in an $A(2.5)$ F2 (light blue).
    \plri{make axis labels bigger}
  }
\end{figure}
\end{example}
%
%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Systems drift and the accumulation of incompatibilities}
Thus far we have shown that many distinct molecular mechanisms can realize identical phenotypes
and that these mechanisms may fail to produce viable hybrids.
Here we discuss the rate at which evolution shifts molecular mechanisms
and ask whether this process is fast enough to be a significant driver of speciation.
To do so, we explore a general quantitative genetic model in which a population drifts stochastically
near a set of equivalent and optimal systems. We work with a population of effective size $N_e$.

>>>>

%Define $V(\tau)$ as the parameterized change of coordinates matrix that preserves $B$ and $C$, with $\tau$ a vector of free parameters. The set of \emph{all} phenotypically invariant (minimal) gene networks is, 
 % \begin{align}
 %   A(\tau) = V(\tau) A(0) V^{-1}(\tau) ,
 % \end{align}
%and a \emph{Linear Evolutionary System} is, 
%\plr{why is this "evolutionary"? It is the thing that we're thinking about evolution of, but that doesn't make it evolutionary.}
Therefore the set of all phenotypically identical (and minimal) systems can be parameterized as, $\mathcal{S}_{\min}(V) = (VAV^{-1}, VB, CV )$,
  %  \begin{align} \label{eq:min_param}
  %    \mathcal{S}(V) := \left\{ \begin{array}{ll} V\dot{\kappa}(t) &= VA V^{-1} \kappa(t) + VB u(t) \\ \phi(t) &= C V^{-1} \kappa(t) \end{array} \right.
  %\end{align}
  where $V$ is any invertible matrix and it's elements are free parameters.

@@

More generally,
we denote by $\calA_n(\Sys_0)$ the set of all $n$-dimensional systems equivalent to $\Sys_0$:
\begin{equation} \label{eqn:equivalence}
  \begin{aligned}
    \calA_n(\Sys_0) 
      &= \left\{
        (A,B,C) : C e^{At} B = C_0 e^{A_0 t} B_0 \; \text{for}\; t \ge 0 
      \right\}  \\
      &= \left\{
        (A,B,C) : C A^r B = C_0 A_0^r B_0 \; \text{for}\; 1 \le r \le n-1 
      \right\} .
  \end{aligned}
\end{equation}
Equivalence of the two characterizations follows from the Cayley-Hamilton theorem.
\plr{To make sense of this we need to say how $C$ and $B$ change with $n$.}
Usually, the dimension $n$ and the reference system $\Sys_0$ is implicit and we write only $\calA$.
Further, we refer to the set of all linear systems with impulse response $h(t)$, regardless of dimension, to be $\allS$,
  \begin{align}
    \allS := \bigcup_{n=\min}^{\infty} \calA_n(\Sys_0)
  \end{align}
  \js{what do you like better for $\allS$? $\mathcal{A}$, $\mathcal{A}_{\infty}$, or $\mathcal{K}_{h(t)}$, or something else -- maybe a modification of $\Sys$? Using a script of ``A'' was more intuitive when we were leavng $B$ and $C$ invariant.}
Two systems, even if in different dimensions, will have identical phenotypes if they are in $\allS$. Further, there is no unique triple $(A,B,C)$ for the mapping $h: u \mapsto \phi$. This implies that the gene regulatory network architecture per phenotype/environment pair is never unique -- that is under these assumptions, \emph{all} gene regulatory networks can drift/rewire, and still conserve the phenotype. This set can be precisley defined and completely parameterized using the \emph{Kalman decomposition} (Appendix \ref{apx:kalman}).

@@

A large, adapted population will move around in $\allS$ under environmental and selective stasis. In the minimal dimension neutral evolution proceeds as a stochastic coordinate transformation on the elements in $\Sys_{\min}$. Even if there is conservation of phenotype through evolutionary time, the molecular mechanism underpinning the phenotype is not constrained and likely will not be conserved.

@@

%\plr{This is not part of the definition of $\mathcal{I}$:}
%\js{should we delete this example ($A(0) \times A(2)$ hybrids)? I like how clean it is -- but I'm not sure it's that important in this context. It's a good proof of principle though.}
%\jss{$F_1$s created by crossing phenotypically equivalent oscillators $A(0)$ and $A(2)$ have a phenotype of $\phi_{F_{1}}(t) = e^{t}$, 
%in contrast to both parents, who have $\phi_{P_{1}}(t) = \phi_{P_{2}}(t) = \sin(t) + \cos(t)$. 
%The hybrid phenotype is significantly different (it does not oscillate and increases infinitely) despite the phenotypic equivalence of the parents.
%%\plr{don't need these to be big equations.}
%%\begin{align*}
%%  \fit \left(\phi_{P_{1}}\right) = \fit \left( \phi_{P_{2}} \right) &= 1 \\
%%  \fit\left(\phi_{F_{1}}\right) &= 0 \\
%%  \mathcal{I} &= 0
%%\end{align*}
%
%%Thus if populations $1$ and $2$ are homogeneous $A(0)$ and $A(2)$, respectively, we say that they are completely incompatible as $\mathcal{I} = 0$.
%
%  \begin{figure}[H]
%    \centering
%    \includegraphics[width=0.5\textwidth, height=0.25\paperheight]{expF1}
%    \caption{$F_1$ hybrid (orange) and parental (blue) phenotypic oscillator dynamics for an $A(0)$ by $A(2)$ cross. The hybrid fails to oscillate and exhibits qualitatively different dynamics.}
%  \end{figure}
%  }
%%     \begin{example}[F1 Reproductive Incompatibility in an Oscillating Gene Network]
%%       DMI examples...
%%       \begin{figure}[H]
%%         \includegraphics[width=0.5\textwidth]{A2_A3_osci_F1_hyb}
%%       \end{figure}
%%     \end{example}
%
%%     \begin{example}[Not all Networks can Host Incompatibilities]
%%       convex sets cant have DMIs
%
%%       \begin{align*}
%%         h(t) = 2 e^{- \theta t}
%%       \end{align*}
%%       Any non-minimal system with rows summing to $\theta$ is PI. Further, these systems are closed under averaging (mating) and row swapping (meiosis), leaving all hybrids optimally fit. The set of gene matrices is affine and therefore convex.  
%%     \end{example}
%
%
%\plri{Missing: a statement about when there are neutral directions, and how many.}
%
%\plr{I think we should first put the section about quantitative drift in, and then the following section that tries to get at real values of parameters.}
%
%
 \begin{example}[All Phenotypically Equivalent Cell Cycle Control Networks] \label{ex:all_osc}
   The set of all two-gene regulatory networks phenotypically equivalent to the cell cycle control network in \ref{ex:oscillator}, where only $A$ can vary, is given by $\Sys (\tau) = (A(\tau), B, C)$, such that, 
    %\begin{align*}
    %  A(\tau) &= \begin{bmatrix} 1 & 0 \\ \tau & 1-\tau \end{bmatrix} \begin{bmatrix} 0 & 1 \\ -1 & 0 \end{bmatrix} \begin{bmatrix} 1 & 0 \\ \frac{\tau}{\tau-1} & \frac{-1}{\tau-1} \end{bmatrix}  \qquad \forall \ \tau \neq 1 %\\
    %\end{align*}
    \begin{align*}
      A(\tau) &= \frac{1}{\tau-1} \begin{bmatrix} \tau & -1 \\ 2 \tau(\tau - 1) + 1 &  -\tau \end{bmatrix} \ \text{for} \ \tau \neq 1
    \end{align*}
    \begin{figure}[H]
    \centering
% Nice tikz!!!
\begin{tikzpicture}
\begin{scope}[every node/.style={circle,thick,draw}]
  \node (A) at (0,0) {$\kappa_{1}$};
    \node[dashed] (B) at (4,0) {$\kappa_{2}$};
    \node[shape=rectangle] (U) at (2,2) {input};
    \node[shape=rectangle] (y) at (2,-2) {output};
\end{scope}

@@

\end{tikzpicture}
      \caption{$\Sys(\tau)$, the set of all phenotype-equivalent cell cycle control networks $\Sys (\tau) = \left\{ \calA_2 : B = [1\ 1]^{T} ,\ C = [1\ 0] \right\}$.}
    \end{figure}
 % \end{example}

@@

   Despite the phenotypic equivalence of all instantiations of $\Sys(\tau)$, the kryptotypes, vary as a function of $\tau$. 

@@

    \caption{Gene-1 dynamics (blue) for both systems $\Sys(0)$ and $\Sys(2)$ are identical, however, $\Sys(0)$ gene-2 dynamics (orange) differ from $\Sys(2)$ (green). Both gene-1 dynamics are given by $\kappa_{1} = \sin t + \cos t$, and gene-2 by $\kappa_{2} = \cos t - \sin t$ ($\Sys(0)$) and $\kappa'_{2} = \cos t + 3 \sin t$ ($\Sys(2)$).}

@@

%\paragraph*{Hybrid phenotypic breakdown between oscillators}
%For example, if the two phenotypically equivalent systems $A(0)$ and $A(2)$ from Example \ref{ex:all_osc} hybridize, their offspring's phenotype will fail to oscillate completely. 
  \begin{figure}[H] 
    \centering
    \includegraphics[width=0.5\textwidth, height=0.125\paperheight]{expF1}
    \caption{\textbf{Hybrid phenotypic breakdown} $F_1$ hybrid (orange) and parental (blue) phenotypic cell-cycle control (oscillator) dynamics for an $\Sys(0)$ by $\Sys(2)$ cross from Example \ref{ex:all_osc}. The $F_1$ hybrid
%$\left(A_{F_1}\!\!=\!\!\tfrac{A\!(\!0\!) \! + \! A\!(\!2\!)}{2}\right)$
gene expression fails to oscillate and instead grows exponentialy
($\phi_{F_0}\! = \! \sin t \! + \! \cos t$, however $\phi_{F_1} \! = \! e^t$).
    }
    \label{fig:expF1}
  \end{figure}
%
%\paragraph{Sexual reproduction and speciation}
%A diploid organism contains two gene network copies: one from each parent. As such, a diploid's system dynamics is computed by averaging the coefficients of both systems. To reproduce, each organism passes on a recombined haploid gene network to its offspring. Haploid gene networks swap system matrix rows randomly between its own two networks.
%Therefore the phenotypic dynamics typical of a first generation ($F_{1}$) cross between two allopatric populations will be determined simply by averaging it's two parental genomes. However, in second generation hybrid ($F_2$) crosses, first new haploid systems will be formed by recombination -- in the process shuffling and combining regulatory coefficients from allopatric populations -- and then brought together to form a diploid.
%
%
  \begin{example}[Hybrid Incompatibility in an Oscillating Gene Network] \label{ex:hybrid_osc}
    Here we compare the phenotypes for $F_2$ hybrids formed by crossing oscillators with different values for $\tau$ from \ref{ex:all_osc}, $\Sys(2)$ with $\Sys(2.01)$, $\Sys(2.1)$, and $\Sys(2.5)$.
    Systems have identical phenotypes, however some of the hybrids exhibit markedly different dynamics.

@@

\begin{figure}[H]
  \centering
  \begin{tabular}{ccc}
    \includegraphics[width=0.25\textwidth, height=0.125\paperheight]{F1_comparison} & 
    \includegraphics[width=0.25\textwidth, height=0.125\paperheight]{F2s_comparison2} &
    \includegraphics[width=0.25\textwidth, height=0.125\paperheight]{figures/quartic_quadratic2}
  \end{tabular}
  \caption{
    $F_1$ (left) and $F_2$ (right) hybrids crossing $\Sys(2)$ with $\Sys(2.01)$ (top), $\Sys(2.1)$ (middle), and $\Sys(2.5)$ (bottom).
    $F_2$ hybrids display more phenotypic divergence than $F_1$s, on average. Further, some $F_2$s completely fail to oscillate, as seen in an $\Sys(2.5)$ $F_2$ (light blue).
    \plri{make axis labels bigger}
  }
\end{figure}
\end{example}
%
%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Systems drift and the accumulation of incompatibilities}
Thus far we have shown that many distinct molecular mechanisms can realize identical phenotypes
and that these mechanisms may fail to produce viable hybrids.
Here we discuss the rate at which evolution shifts molecular mechanisms
and ask whether this process is fast enough to be a significant driver of speciation.
To do so, we explore a general quantitative genetic model in which a population drifts stochastically
near a set of equivalent and optimal systems. We work with a population of effective size $N_e$.
